---
title: 'STAT3215 Final Project: Analysis of Aircraft Flight Delay'
author: "Charitarth Chugh (charitarth@uconn.edu), Leon Nguyen (leon.nguyen@uconn.edu)"
date: "12/08/23"
output: pdf_document
---
```{r setup}
knitr::opts_chunk$set(echo = TRUE)
library(plyr)
library(tidyverse)
library(psych)
```

## Section 1. Introduction

In this report, we will reproduce, analyze, and enhance findings from a research paper written by Kalliguddi and Leboulleuc, which seeks to investigate factors contributing to aircraft flight delays within the airline industry. The paper aims to analyze the on-time performance of domestic flights for the year 2016 and develop a better predictive model to forecast flight delay. This is important because flight delays can incur significant economic losses, damage reputability for airlines, and create inconvenience for passengers.

The remainder of the report is divided into four sections as follows. Section 2 discusses the data description and data source. Section 3 describes the data analysis and predictive modeling techniques used in the study, delving into multiple linear regression, assumption validation, and other miscellaneous techniques. Section 4 describes our additional subanalyses and enhancement to the initial paper. Finally, section 5 provides conclusions and further discussion of the initial study and our enhancement.



```{r}
# Getting the compressed data, reading it, and filtering for missing values.

data.dir <- "./Chugh-Charitarth-Data"
df_list <- list()
for (i in list.files(data.dir)) {
  if (i != "readme.html") {
    d <- read_csv(file.path(data.dir, i))
    d <- d %>% filter(DepDelay > 15)
    df_list[[i]] <- d
    sum_rows <- sum_rows + nrow(d)
  }
}
df <- ldply(df_list, rbind) # merge all the data frames into one.
df <- df %>% select(
  DepDelay, CarrierDelay, TaxiIn, TaxiOut,
  Distance, WeatherDelay, NASDelay,
  SecurityDelay, LateAircraftDelay
)
df <- filter(df, !if_any(everything(), is.na))
```

```{r}
summary(df)
```
```{r}
head(df)
```